Forecasting climate change impacts on plant population dynamics at large spatial extents: a test case with sagebrush (\emph{Artemisia}) species

### 2015-06-29 docs/ms_sageAbundance.Rmd
updates ms with kernel rationale

%----------------------------------------------------
\section{\large Introduction}
%----------------------------------------------------
Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology. The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions. However, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale. Local-scale demographic data make building population projection models an easy task (Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions. Also, given the costliness (in time and money) of collecting detailed demographic data, it is unlikely that “enough” local-scale data can be collected to accurately project large-scale population dynamics.

@@

\section{Materials and Methods}
\subsection{Data: Remotely-sensed time series}
To show our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure [fig:Location-of-the]). Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming. The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65. To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. This resulted in an annual time series of sagebrush cover at 30 m resolution from 1984 to 2011. In this dataset, values represent the percentage of a 30 $\times$ 30 m pixel covered by sagebrush. For our purposes we use two 5 $\times$ 5 km subsets (study areas 1 and 2), totaling 27,948 pixels each. Thus, each full dataset contains 754,596 observations (27 years $\times$ 27,948 pixels).

\begin{figure}[h]
\center{}\includegraphics[width=1\textwidth]{/Users/atredenn/Dropbox/sagebrush_class_2013/Wyoming/figures/StudyAreasMap_1and2}

\protect\caption{\label{fig:Location-of-the}Location of the study area in southwestern
Wyoming and a snapshot of the percent cover data in 1984 for each
site (a and b). Green shading on the map shows projected distribution
of sagebrush from Kuchler (19xx). Latitude and longitude are in decimal
degrees.}

\end{figure}

\subsection{Data: Climate covariates}
Our approach is based on modeling how changes in plant populations relate to climate variables. We a priori narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model future projections. In all vital rate regressions we focus on the demographic processes driving changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t}, and how these processes are mediated by climate. The four climate variables in our vital rate regressions are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table [tab:Definition-of-climate]). Climate data to parameterize vital rate regressions (`observed' climate) came from the PRISM climate database. Climate data were extracted for the centroid our study area.

Climate data used to inform model projections of future sagebrush
cover came from average outputs of all models used in the most recent
IPCC report, the Coupled Model Intercomparison Project 5 (CMIP5; http://cmip-pcmdi.llnl.gov/cmip5/).
We downloaded model estimates of historic climate (1950-2000) and
projected climate (2050-2100) for the centroid of our study area.
We wanted to keep the temporal variability from observed climate consistent,
so our aim was to adjust observed climate according to projected changes.
For temperature we simply subtracted the average historic temperature
from the average projected temperature. For precipitation we calculated
the proportional change: we subtracted average projected precipitation
from average historic precipitation, and then divided by average historic
precipitation. We did this for three ``Representative Concentration
Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5. RCPs are described here,
http://tntcat.iiasa.ac.at/RcpDb/.
Basically they correspond to stabilization of radiative forcing before
2100, after 2100, and increasing greenhouse gas emissions, respectively.
We implemented projected temperature and precipitation changes in
the model by perturbing observed annual values by the values in Table
\ref{tab:Changes-in-climate}.

\begin{table}[h]
\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
average projections.}

\begin{tabular}{llllll}
\hline
 & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
\hline
RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
\hline
\end{tabular}
\end{table}

\subsection{Spatio-temporal model for sagebrush cover}
We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects. Since the spatial and temporal effects are additive, we are not considering space-time interactions. Our model aims to relate observed sagebrush cover, $C_{s,t}$, at time \emph{t} and spatial location \emph{s} to an underlying process $\mu_{s,t}$ that is defined over all spatial locations and observation times. For statistical convenience, and with little added bias, we consider discrete values of sagebrush percent cover as counts, giving us the formulation

@@

\noindent which states that counts (integer percent cover) are Poisson distributed with a log link function on abundance intensity ($\boldsymbol{\lambda}$). We assume the underlying process, $\boldsymbol{\mu}$, is a function of climate covariates (\textbf{X}), lagged sagebrush abundace (a temporal process), and a purely spatial random effect ($\boldsymbol{\eta}$):

@@

Fitting the spatial random effect ($\eta$) is computationally demanding when the observations come from 10s of 1,000s locations. The dimensionality is just too large. To overcome these computational constraints we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation. A common strategy is to express high dimensional effects, like $\boldysmbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$. We can then define the spatial effect as

@@

\noindent In this case, $\boldsymbol{\alpha}$ is a $m \times 1$ vector of reduced spatial random effects, and \textbf{K} is a $S \times m$ matrix that maps the reduced effects to the full \emph{S}-dimensional space, where \emph{S} is the total number of observed locations. Thus, we are able to reduce the effective number of parameters from ${S} \rightarrow m$.

Obviously, the last remaining obstacle is to parameterize \textbf{K}. Here we use kernel convolution to interpolate the spatial random effect between \emph{m} ``knots'' that are nonrandomly distributed across the space of our study area. These means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots. We use a simple Gaussian kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are

\begin{align}
K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \phi}),
\end{align}

\noindent where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth. Based on residual spatial structure we found after fitting a simple model without spatial effects, we set $\phi$ to about one-third of the range of spatial dependence evident in the residual variograms -- 250 meters in this case. In the end, the simplest way to understand our dimension reduction strategy is to consider $\alpha$s as weights for each knot.

\subsection{Computing and model estimation}

\end{document}

>>>>

Introduction
-------------

Forecasting the impacts of climate change on plant populations and communities is a central challenge of modern ecology. The challenge lies in faithfully representing population dynamics at spatial and temporal scales relevant to policy and management decisions. However, almost every study of plant population dynamics relies on demographic observations recorded at the meter to sub-meter scale. Local-scale demographic data make building population projection models an easy task (Ellner et al., 2006; Rees et al., 2009; Adler et al., 2012), but it is very difficult to extrapolate small-scale studies to large spatial extents with any certainty because the data likely only represent a small subset of parameter space and environmental conditions. Also, given the costliness (in time and money) of collecting detailed demographic data, it is unlikely that “enough” local-scale data can be collected to accurately project large-scale population dynamics.

@@

Materials and Methods
---------------------
### Data: Remotely-sensed time series
To show our pixel-based modeling approach we use a subset of a remotely-sensed time series of sagebrush (\emph{Artemisia} spp.) shrub percent cover in Wyoming (Homer et al., 2012) (Figure [fig:Location-of-the]). Sagebrush percent cover was estimated using a regression tree to relate ground reflectances retrieved by three sources of optical imagery (QuickBird, Landsat, and AWiFS) to 1,780 field observations of sagebrush cover distributed across Wyoming. The regression tree model was further validated using another 500 field observations. For Wyoming sagebrush, the model achieved an $R^{2}$ = 0.65. To hind-cast sagebrush cover the regression tree model was applied to historical remote sensing images. This resulted in an annual time series of sagebrush cover at 30 m resolution from 1984 to 2011. In this dataset, values represent the percentage of a 30 $\times$ 30 m pixel covered by sagebrush. For our purposes we use two 5 $\times$ 5 km subsets (study areas 1 and 2), totaling 27,948 pixels each. Thus, each full dataset contains 754,596 observations (27 years $\times$ 27,948 pixels).

### Data: Climate covariates
Our approach is based on modeling how changes in plant populations relate to climate variables. We a priori narrowed our focus to climate covariates we know are important for sagebrush and that could be easily calculated from general circulation model future projections. In all vital rate regressions we focus on the demographic processes driving changes in percent sagebrush cover in year \emph{t}-1 to year \emph{t}, and how these processes are mediated by climate. The four climate variables in our vital rate regressions are: (1) year \emph{t} fall through summer precipitation, (2) year \emph{t}-1 fall through summer precipitation, (3) year \emph{t} average spring temperature, and (4) year \emph{t}-1 average spring temperature (Table [tab:Definition-of-climate]). Climate data to parameterize vital rate regressions (`observed' climate) came from the PRISM climate database. Climate data were extracted for the centroid our study area.

Climate data used to inform model projections of future sagebrush
cover came from average outputs of all models used in the most recent
IPCC report, the Coupled Model Intercomparison Project 5 (CMIP5; http://cmip-pcmdi.llnl.gov/cmip5/).
We downloaded model estimates of historic climate (1950-2000) and
projected climate (2050-2100) for the centroid of our study area.
We wanted to keep the temporal variability from observed climate consistent,
so our aim was to adjust observed climate according to projected changes.
For temperature we simply subtracted the average historic temperature
from the average projected temperature. For precipitation we calculated
the proportional change: we subtracted average projected precipitation
from average historic precipitation, and then divided by average historic
precipitation. We did this for three ``Representative Concentration
Pathways'': RCP 4.5, RCP 6.0, and RCP 8.5. RCPs are described here,
http://tntcat.iiasa.ac.at/RcpDb/.
Basically they correspond to stabilization of radiative forcing before
2100, after 2100, and increasing greenhouse gas emissions, respectively.
We implemented projected temperature and precipitation changes in
the model by perturbing observed annual values by the values in Table #.

\begin{table}[h]
\protect\caption{\label{tab:Changes-in-climate}Changes in climate variables from CMIP5
average projections.}
\begin{tabular}{llllll}
\hline
 & \multicolumn{2}{c}{Study Area 1} &  & \multicolumn{2}{c}{Study Area 2}\tabularnewline 
Emissions Scenario & $\Delta$ temperature & $\Delta$ precipitation &  & $\Delta$ temperature & $\Delta$ precipitation\tabularnewline
\hline
RCP 4.5 & 2.98\textdegree & 8.94\% &  & 3.04\textdegree & 7.42\%\tabularnewline
RCP 6.0 & 3.13\textdegree & 8.64\% &  & 3.23\textdegree & 6.72\%\tabularnewline
RCP 8.5 & 4.79\textdegree & 11.0\% &  & 4.90\textdegree & 8.81\%\tabularnewline
\hline
\end{tabular}
\end{table}

### Additive spatio-temporal model for sagebrush cover
We use a descriptive model for sagebrush cover that includes additive spatial and temporal effects. Since the spatial and temporal effects are additive, we are not considering space-time interactions. Our model aims to relate observed sagebrush cover, $C_{s,t}$, at time \emph{t} and spatial location \emph{s} to an underlying process $\mu_{s,t}$ that is defined over all spatial locations and observation times. For statistical convenience, we consider discrete values of sagebrush percent cover as counts, giving us the formulation

@@

which states that counts (integer percent cover) are Poisson distributed with a log link function on abundance intensity ($\boldsymbol{\lambda}$). We assume the underlying process, $\boldsymbol{\mu}$, is a function of climate covariates (\textbf{X}), lagged sagebrush abundace (a temporal process), and a purely spatial random effect ($\boldsymbol{\eta}$):

@@

Fitting the spatial random effect ($\eta$) is computationally demanding when the observations come from 10s of 1,000s locations. The dimensionality is just too large. To overcome these computational constraints we employed a dimension reduction strategy to reduce the number of parameters that need to be estimated to account for spatial variation. A common strategy is to express high dimensional effects, like $\boldsymbol{\eta}$, as the product of an expansion matrix, \textbf{K}, and a smaller parameter vector, $\boldsymbol{\alpha}$. We can then define the spatial effect as

@@

In this case, $\boldsymbol{\alpha}$ is a $m \times 1$ vector of reduced spatial random effects, and \textbf{K} is a $S \times m$ matrix that maps the reduced effects to the full \emph{S}-dimensional space, where \emph{S} is the total number of observed locations. Thus, we are able to reduce the effective number of parameters from ${S} \rightarrow m$.

Obviously, the last remaining obstacle is to parameterize \textbf{K}. Here we use kernel convolution to interpolate the spatial random effect between \emph{m} "knots" that are nonrandomly distributed across the space of our study area. These means we are modeling spatial random effects at the knot level, and we use \textbf{K} to interpolate those effects between knots. We use a simple Gaussian kernel density to define the distance-decay function around the knots (\textbf{w}), such that the entries of \textbf{K} are

\begin{align}
K_{s,m} &= w_{s,m}/\sum_{s} w_{s,m}, \text{where} \\
w_{s,m} &= (2 \pi \sigma^{2})^{-1} \text{exp}(\frac{-d_{s,m}}{2 \sigma^{2}}),
\end{align}

where $d_{s,m}$ is the Euclidean distance between the centroid of sample cell \emph{s} and the location of knot \emph{m}, and $\sigma$ is the kernel bandwidth. It is possible through exhaustive model selection and fitting to determine the optimal form of the kernel and to estimate optimal values for $\sigma$. However, given the relative largeness of our dataset and computational limitations, we decided to to create kernels around eight knots and used a rule-of-thumb to set $\sigma$ equal to the distance between nearest neighbor knots (Higdon 2002, Conn et al. 2014).

### Computing and model estimation
